---
title: "Untitled"
author: "Janish Parikh"
date: "12/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(GGally)
library(lubridate)
library(caret)
library(gbm)
library(tidyverse)
library(xgboost)
library(caret)
```

```{r, cache=TRUE}
citibike_2019 <- read.csv("~/citibike_2019.csv")
citibike_2019<-citibike_2019[sample(nrow(citibike_2019),nrow(citibike_2019)/10),]
citibike_2019 <- citibike_2019 %>% mutate(tripduration = tripduration /60)
```

```{r}
is_weekday = function(timestamp){
  lubridate::wday(timestamp, week_start = 1) < 6
}
```

Feature Extraction
```{r, cache=TRUE}
citibike_2019$Hour<-cut(hour(citibike_2019$starttime), breaks = c(-1, 5, 12, 16, 20, 24),labels = c("Night", "Morning", "Afternoon", "Evening", "Night"))
citibike_2019$dayofweek <- as.factor(wday(citibike_2019$starttime))
citibike_2019$weekday<-as.factor(as.numeric(sapply(citibike_2019$starttime, is_weekday)))
```

Trip Duration
Max trip duration 3 hours 
```{r, cache=TRUE}
citibike_2019<-citibike_2019[citibike_2019$tripduration<=10800,]
```

AGE
```{r, cache=TRUE}
citibike_2019$age<- 2019 - citibike_2019$birth.year
summary(citibike_2019$age)
citibike_2019<-citibike_2019[citibike_2019$age<=90,]
```

Distance
```{r, cache=TRUE}
library(geosphere)
citibike_2019$distance <- distHaversine(citibike_2019[,6:7], citibike_2019[,10:11])
citibike_2019$dist.lat <- abs((citibike_2019$start.station.latitude - citibike_2019$end.station.latitude))
citibike_2019$dist.long <- abs((citibike_2019$start.station.longitude - citibike_2019$end.station.longitude))
summary(citibike_2019$distance)
```

Convert into factor variables
```{r}
citibike_2019$usertype<-as.factor(citibike_2019$usertype)
citibike_2019$gender<-as.factor(citibike_2019$gender)
```

Weather Data
```{r, cache=TRUE}
weather <- read.csv("~/Downloads/2812766.csv",header=FALSE)
names(weather)<-weather[1,]
weather<-weather[-1,]
weather<-subset(weather[weather$NAME == "NY CITY CENTRAL PARK, NY US" & year(weather$DATE) == 2019,],)
weather<-subset(weather, select = -c(STATION,NAME,LATITUDE,LONGITUDE,ELEVATION, AWND_ATTRIBUTES, DAPR_ATTRIBUTES, MDPR_ATTRIBUTES,PGTM_ATTRIBUTES,PRCP_ATTRIBUTES,PSUN_ATTRIBUTES,SNOW_ATTRIBUTES,SNWD_ATTRIBUTES,TAVG_ATTRIBUTES,TMAX_ATTRIBUTES,TMIN_ATTRIBUTES,TOBS_ATTRIBUTES,TSUN_ATTRIBUTES,WDF2_ATTRIBUTES,WDF5_ATTRIBUTES,WESD_ATTRIBUTES,WESF_ATTRIBUTES,WSF2_ATTRIBUTES,WSF5_ATTRIBUTES,WT01_ATTRIBUTES,WT02_ATTRIBUTES,WT03_ATTRIBUTES,WT04_ATTRIBUTES,WT05_ATTRIBUTES, WT06_ATTRIBUTES, WT08_ATTRIBUTES,WT09_ATTRIBUTES,WT11_ATTRIBUTES))
weather<-weather %>%mutate(across(c(where(is.character), -DATE), as.numeric))%>%mutate_all(~replace_na(., 0))
weather$DATE<-as.Date(weather$DATE)
head(weather)
```
Join Citibike data with Weather Data
```{r, cache=TRUE}
citibike_2019$start_date<-date(citibike_2019$starttime)
citibike_2019<-merge(citibike_2019, weather, by.x = "start_date", by.y = "DATE")
names(citibike_2019)
```

```{r}
colstouse <-setdiff(names(citibike_2019),c("start_date","starttime","stoptime","start.station.name","end.station.name"))
train_data<-citibike_2019[ , which(names(citibike_2019) %in% colstouse)]
```
Find important features
```{r, cache=TRUE}
library(Boruta)
boruta.train <- Boruta(tripduration~., data = test, doTrace = 2)
print(boruta.train)
```

```{r}
plot(boruta.train, xlab = "", xaxt = "n")
lz<-lapply(1:ncol(boruta.train$ImpHistory),function(i)
boruta.train$ImpHistory[is.finite(boruta.train$ImpHistory[,i]),i])
ames(lz) <- colnames(boruta.train$ImpHistory)
Labels <- sort(sapply(lz,median))
axis(side = 1,las=2,labels = names(Labels),
at = 1:ncol(boruta.train$ImpHistory), cex.axis = 0.7)
```

```{r, fig.align='center',fig.width=20, fig.height=20, dpi= 2000}
ggcorr(train_data, name = "Correlation", label = TRUE, alpha = TRUE, palette = "PuOr") + ggtitle("Correlation Matrix Plot") + theme_light()
```
```{r}
colstouse<-c("tripduration","gender","age","distance","dist.lat","dist.long","usertype","dayofweek","AWND","PRCP","SNOW","SNWWD","TMAX","TMIN","WDF2","WDF5","WSF5","WT01","WT02")
```

```{r}
train_data<-citibike_2019[ , which(names(citibike_2019) %in% colstouse)]
linear_model<-lm(tripduration~., data = train_data)
```

```{r}
summary(linear_model)
```

```{r}
test <- read.csv("~/Downloads/citybike.test10000Awithduration.csv")
test<-test[test$tripduration<=3600,]
test$Day <- as.numeric(substr(test$starttime,9,10))
test$Year<-year(test$starttime)
test$Hour <- as.numeric(format(as.POSIXct(test$starttime), format = "%H"))
test$dayofweek <- weekdays(as.Date(test$starttime))
test$start_date <- date(test$starttime)
test$Age<- test$Year - test$birth.year
test$weekend<-as.numeric(sapply(test$starttime,is_weekday))
head(subset(test, select = c(Day, Hour,dayofweek,Year, Age, birth.year)))
```

```{r}
test$distance <- distHaversine(test[,4:5], test[,8:9])
test$distance.lat <- abs((test$start.station.latitude - test$end.station.latitude))
test$distance.long <- abs((test$start.station.longitude - test$end.station.longitude))
head(subset(test, select = c(distance,distance.lat,distance.long)))
```

```{r}
test<-merge(test, weather, by.x = "start_date", by.y = "DATE")
test2<-subset(test, select = c(weekend, usertype, gender, Day, Hour,dayofweek,Age,distance,birth.year, distance.lat,  distance.long,PRCP, TMAX, TMIN ))
```

```{r}
y_train <- train1$tripduration
predicted<-predict(linear_model, train1)
residuals = y_test - predicted
RMSE = sqrt(mean(residuals^2))
cat('The root mean square error of the train data is ', round(RMSE,3),'\n')

y_test <- test$tripduration
predicted<-predict(linear_model, test)
residuals = y_test - predicted
RMSE = sqrt(mean(residuals^2))
cat('The root mean square error of the test data is ', round(RMSE,3),'\n')
```


```{r}
y_test_mean = mean(y_test)
# Calculate total sum of squares
tss =  sum((y_test - y_test_mean)^2 )
# Calculate residual sum of squares
rss =  sum(residuals^2)
# Calculate R-squared
rsq  =  1 - (rss/tss)
cat('The R-square of the test data is ', abs(round(rsq,3)), '\n')
```

```{r}
my_data = as.data.frame(cbind(predicted = predicted,observed = y_test))
# Plot predictions vs test data
ggplot(my_data,aes(predicted, observed)) + geom_point(color = "darkred", alpha = 0.5) + 
    geom_smooth(method=lm)+ ggtitle('Linear Regression ') + ggtitle("Linear Model: Prediction vs Test Data") +
      xlab("Predecited Trip Duration ") + ylab("Observed Trip Duration") + 
        theme(plot.title = element_text(size=16,hjust = 0.5),
         axis.text.y = element_text(size=12), axis.text.x = element_text(size=12,hjust=.5),
         axis.title.x = element_text(size=14), axis.title.y = element_text(size=14))
```

# Gradient Boosted Model

```{r}
colstouse<-c("tripduration","gender","age","distance","dist.lat","dist.long","usertype","dayofweek","AWND","PRCP","SNOW","SNWWD","TMAX","TMIN","WDF2","WDF5","WSF5","WT01","WT02")
```

```{r}
library('gbm')
fitControl <- trainControl(
                           method = "repeatedcv",
                           number = 10,
                           repeats = 10)

gbmmodel <- train(tripduration ~ ., 
                 data = train_data,
                 distribution = "gaussian",
                 method = "gbm", 
                 trControl = fitControl,
                 verbose = F)
```

```{r}
print(gbmmodel)
summary(gbmmodel)
names(test1)
```
```{r}
set.seed(123)
# train GBM model
gbm.fit <- gbm(
  formula = tripduration ~ .,
  distribution = "gaussian",
  data = train_data,
  n.trees = 10000,
  interaction.depth = 1,
  shrinkage = 0.001,
  cv.folds = 5,
  n.cores = NULL, # will use all cores by default
  verbose = FALSE
  )  
```

```{r}
summary(gbm.fit)
print(gbm.fit)
```

# XGBoost
```{r}
train2<-train_data
train2$gender<-as.numeric(as.factor(train2$gender))
train2$usertype<-as.numeric(as.factor(train2$usertype))
train2$dayofweek<-as.numeric(as.factor(train2$dayofweek))
summary(train2)
X_train = as.matrix(subset(train2, select = -c(tripduration)))
y_train = train2$tripduration
```

```{r}
xgb_trcontrol = trainControl(
  method = "cv",
  number = 10,  
  allowParallel = TRUE,
  verboseIter = T,
  returnData = FALSE
)
```

```{r}
xgbGrid <- expand.grid(nrounds = 200,
                       max_depth = 20,
                       colsample_bytree = 0.9,
                       eta = 0.1,
                       gamma=0,
                       min_child_weight = 1,
                       subsample = 1
                      )
```

```{r, warning=FALSE}
set.seed(0) 
xgb_model = train(
  X_train, y_train,  
  trControl = xgb_trcontrol,
  tuneGrid = xgbGrid,
  method = "xgbTree"
)
```
```{r}
summary(xgb_model)
print(xgb_model)
```

```{r}
test2$usertype<-as.numeric(as.factor(test2$usertype))
test2$dayofweek<-as.numeric(as.factor(test2$dayofweek))
X_test<-as.matrix(subset(test2))
head(X_test)
Y_test<-test2$tripduration
```

```{r}
y_test<-train_data$tripduration
predicted = predict(gbmmodel,train_data)
residuals = y_test - predicted
RMSE = sqrt(mean(residuals^2))
cat('The root mean square error of the test data is ', round(RMSE,3),'\n')
```

```{r}
y_test_mean = mean(y_test)
# Calculate total sum of squares
tss =  sum((y_test - y_test_mean)^2 )
# Calculate residual sum of squares
rss =  sum(residuals^2)
# Calculate R-squared
rsq  =  1 - (rss/tss)
cat('The R-square of the test data is ', round(rsq,3), '\n')
```

```{r}
xgb_model
importance_matrix <- xgb.importance(model = xgb_model)
importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix, xlab = "Feature Importance")
```

```{r}
options(repr.plot.width=8, repr.plot.height=4)
my_data = as.data.frame(cbind(predicted = predicted,
                            observed = y_test))
# Plot predictions vs test data
ggplot(my_data,aes(predicted, observed)) + geom_point(color = "darkred", alpha = 0.5) + 
    geom_smooth(method=lm)+ ggtitle('Linear Regression ') + ggtitle("Extreme Gradient Boosting: Prediction vs Test Data") +
      xlab("Predecited Power Output ") + ylab("Observed Power Output") + 
        theme(plot.title = element_text(color="darkgreen",size=16,hjust = 0.5),
         axis.text.y = element_text(size=12), axis.text.x = element_text(size=12,hjust=.5),
         axis.title.x = element_text(size=14), axis.title.y = element_text(size=14))
```

```{r}
pred_xgb <- predict(m1_xgb, as.matrix(test[, 11:19]))
pred<-as.data.frame(x=predicted)
pred$ID<-c(1:20000)
colnames(pred) <- c("Predicted","ID")
pred<-subset(pred, select = c(ID, Predicted))
write.csv(pred, file = "valhalla5.csv", row.names = FALSE)

pred1 <- read.csv("predictions.csv")
pred1 <- pred1[1:20000,]
colnames(pred1) <- c("Predicted","ID")
pred1<-subset(pred1, select = c(ID, Predicted))
pred1 <- pred1[1:20000,]
write.csv(pred1, file = "valhalla2.csv", row.names = FALSE)
```

function(predicted,actual){
  #MAE
  mae <- mean(abs(actual-predicted))
  #MSE
  mse <- mean((actual-predicted)^2)
  #RMSE
  rmse <- sqrt(mean((actual-predicted)^2))
  #MAPE
  mape <- mean(abs((actual-predicted)/actual))
  errors <- c(mae,mse,rmse,mape)
  names(errors) <- c("mae","mse","rmse","mape")
  return(errors)
}
